Generation of Variation and a Modified Mean Fitness Principle: \\ Necessity is the Mother of Genetic Invention

### 2018-02-09 ms.tex
reformat margins, font size, equation overflows
removed titlesec, messed up nameref
MWF and LA revisions for TPB

\documentclass[9pt, a4paper, twocolumn]{extarticle}
\usepackage[margin=1in]{geometry} % See geometry.pdf to learn the layout options.
\geometry{a4paper}
%\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx} % Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
% TeX will automatically convert eps --> pdf in pdflatex

@@

% messes up nameref
%\usepackage{titlesec}
%\titleformat{\subsection}[runin]
%  {\bf}{\thesection}{1em}{}

@@

% Title page
\title{Generation of Variation and Mean Fitness Increase: \\ Necessity is the Mother of Genetic Invention}
%\title{Conditional Generation of Variation and Mean Fitness Increase: \\ Necessity is the Mother of Genetic Invention}
% \title{A General Principle for the Evolution of Fitness-Associated Generation of Variation}

@@

Another departure from the \emph{linear variation} assumption of the \emph{reduction principle} for mutation rates involves a mechanism by which the mutation rate increases in individuals of low fitness -- a mechanism first observed in stressed bacteria~\citep{Foster2007}, although not in a constant environment.
\citet{Ram2012} demonstrated that even in a constant environment, increasing the mutation rate of individuals with below-average fitness increases the population mean fitness, rather than decreases it.
Their analysis assumed  infinite population size and fitness  determined by the number of mutant alleles accumulated in the genotype.
In their models, the only departure from the \emph{reduction principle} conditions was the unequal scaling of mutation probabilities between different genotypes introduced by the correlation between the mutation rate and fitness.
A similar result has been demonstrated for conditional dispersal~\citep[Th.~39]{Altenberg2012b}, fitness-associated recombination~\citep{Hadany2003a} and for condition-dependent sexual reproduction~\citep{Hadany2007a}, and empirical evidence suggests these mechanisms are common in nature~\citep{Ram2016}.

@@

\paragraph*{General model.}\label{sec:general_model}
Consider a large  population with an arbitrary set of types $A_1, A_2, \ldots, A_n$.
The frequency and fitness of individuals of type $A_k$ are $f_k$ and $w_k$, respectively.
The probability that an individual of type $A_k$ will transition to some other type is $C_k$, and given a transition occurs, the probability that it will transition to type $A_j$ is $M_{j,k}$.
Therefore, the change in the frequencies of type $A_k$ is described by the transformation $f \to f'$: 
\begin{equation}
\bar{w} f'_k = (1-C_k) w_k f_k + \sum_{j=1}^{n}{C_j M_{k,j} w_j f_j},
\label{eq:model_sum}
\end{equation}
or in matrix form
\begin{equation}
\bar{w} f' = (\ci - C + MC)D f,
\label{eq:model}
\end{equation}
where $f=(f_1, f_2, \ldots, f_n)$ is a frequency vector with $f_k\ge 0$ and  $\sum_{k=1}^n{f_k} = 1$;
$D$ is a positive diagonal matrix with entries $w_k$ such that $w_k \ne w_j$ for some $k \ne j$; 
$C$ is a positive diagonal  matrix with entries $C_k$;
$M$ is a primitive column-stochastic matrix: $M_{j,k} \ge 0$ for all $j,k$, $\sum_{j=1}^n {M_{j,k}} = 1$ for all $k$, and $(M^{\ell})_{j,k} > 0$ for all $j,k$ for some positive integer $\ell$ (the latter is the condition for primitivity~\citep[see][Appendix 3]{Otto2007});
$\ci$ is the $n\times n$ identity matrix;
and $\bar{w}$ is the normalizing factor such that $\sum_{k=1}^n{f'_k}=1$ and is equal to the population mean fitness $\bar w=\sum_{k=1}^n f_kw_k$.

@@

Transition between types is determined by a combination of two effects:
(i) the probability of transitioning \emph{out} of type $A_k$ is determined by $C_k$;
(ii) given a transition out of type $A_k$, the distribution of the destination types $A_i$ is given by $M_{i,k}$ (note the index order).
Importantly, different types can have different rates.
That is, $C_i \ne C_j$ for some $i,j$. The case $C_i = C_j$ for all $i,j$ is covered by the \emph{reduction principle}~\citep[see][]{Altenberg2017}.

@@

This model is a special case of the general model (Eq.~\ref{eq:model}) where
\begin{equation} 
{M} = \frac{1}{n-1} \begin{bmatrix}
0 & 1 & 1 & \ldots & 1 \\
1 & 0 & 1 & \ldots & \vdots \\
1 & 1 & 0 & \ldots & \vdots \\
\vdots & \vdots & \vdots & \ddots & \vdots \\
1 & \ldots & \ldots & \ldots & 0
\end{bmatrix},
\end{equation}
with zeros on the diagonal and $\frac{1}{n-1}$ elsewhere.
Note that here ${M}$ is irreducible and primitive.

% Mutation model 2
\paragraph*{Mutation model 2.}\label{sec:mut_model_2}

@@

Therefore, setting 
\begin{equation}
{M} = \begin{bmatrix}
1-\delta & \beta & 0 &  \cdots & 0\\
\delta & 1-\delta-\beta & \beta &  \ddots & 0\\
0 & \delta & 1-\delta-\beta & \ddots & 0 \\
\vdots & \ddots & \ddots & \ddots & \beta \\
0 & 0 & 0 & \delta & 1-\beta
\end{bmatrix},
\end{equation}
with the beneficial, neutral, and deleterious mutation probabilities on the three main diagonals and zeros elsewhere, 
Eq.~\ref{eq:mutation_model_2} can be viewed as a special case of Eq.~\ref{eq:model}.
Here, too, ${M}$ is irreducible and primitive as long as $\delta, \beta > 0$.

@@

After reproduction, offspring of individuals from deme $A_k$ return to their parental deme with probability $1-C_k$, or migrate to a different deme $A_j$ with probability $C_k M_{j,k}$, where the matrix $M$ is primitive, $C_k > 0$, $M_{j,k} \ge 0$, and $\sum_{j=1}^{n}{M_{j,k}} = 1$ for all $k=1, \ldots, n$.
Therefore, $1-C_k$ can be interpreted as a \emph{homing rate}.

@@

For example, in the case of \emph{symmetric individual learning}~\citep{Borenstein2008}, learning is parameterized by its breadth of exploration $b$ and the mature phenotype $j$ is randomly drawn from $2b+1$ phenotypes symmetrically and uniformly distributed around the initial phenotype $k$, with the limitation that any "spillover" of phenotypes below $1$ or above $n$ is "absorbed" by those boundaries.
This "absorption" ensures $M$ is column-stochastic. 
In other words, given initial phenotype $k$, the probability of maturation to phenotype $j$, where $k-b \le j \le k+b$ is $1/(2b+1)$, but any phenotype $j<1$ actually becomes $j=1$ and any phenotype $j>n$ actually becomes $j=n$.
The probability for maturation to other phenotypes is $0$.

@@

% Mean fitness principle
\paragraph*{Mean fitness principle.}

We first focus on the stable population mean fitness.
We  show that if the transition rate from types with below-average fitness increases, then the stable population mean fitness increases, too.

@@

The following result constitutes a \emph{mean fitness principle} for the
sensitivity of the equilibrium mean fitness $\hat{\bar w}$ to changes
in $C_k$, the probability of transition from $A_k$.
\medskip

\begin{result}[Mean fitness principle]\label{result:mfp}
Let $\hat{\bar w}$ be the leading eigenvalue of $(I-C+MC)D$,
and $\hat u$ and $\hat v$ be the corresponding positive left and right eigenvectors, such that $\sum_{k=1}^{n}\hat v_k=1$ and $\sum_{k=1}^{n}\hat u_k\hat v_k=1$.
Then,
\begin{equation}\label{eq:mfp}
\frac{\partial \hat{\bar w}}{\partial C_k} = 
\frac{\hat u_k \hat v_k}{C_k} (\hat{\bar w} - w_k),
\end{equation}
or in simpler terms,
\begin{equation}\label{eq:mfp_sign}
sign\frac{\partial \hat{\bar w}}{\partial C_k} = 
sign(\hat{\bar w} - w_k).
\end{equation}

@@

\begin{proof}
Using the formula in~\citet{Caswell1978} (see eq.~\ref{eq:Caswells_formula} in~\nameref{sec:AppA}),
\begin{equation}
\frac{\partial \hat{\bar w}}{\partial C_k} = 
\hat u\tr \frac{\partial (I-C+MC)D}{\partial C_k} \hat v.
\end{equation}

@@

Then,
\begin{equation}\label{eq:mfp_intermediate}
\begin{aligned}
\hat u\tr \frac{\partial (I-C+MC)D}{\partial C_k} \hat v = 
\hat u\tr (0 - Z_k + M Z_k)D \hat v = \\
-\hat v_k \hat u_k w_k + \hat v_k w_k \hat u\tr[M]_k = \\
\hat v_k w_k (\hat u\tr[M]_k - \hat u_k). 
\end{aligned}
\end{equation}

@@

% Multiple types
\paragraph*{}
We turn our attention to the case where the transition rates from a subset $K$ of the types are equal, that is, $C_j \equiv C_i$ for $i,j \in K$.
In this case, Eq.~\ref{eq:mfp} leads directly to the following by setting $\tau \coloneqq C_i$.

\begin{corollary}
The sensitivity of the stable population mean fitness to change in the rate of transition $\tau$ from types in $K$ is
\begin{equation}\label{eq:mfp_K}
\begin{aligned}
\frac{\partial \hat{\bar w}}{\partial \tau} = 
\hat u\tr \Big( \sum_{k \in K}{\frac{\partial (I - C + MC)D}{\partial \tau}} \Big) \hat v = \\
\frac{1}{\tau} \sum_{k \in K}{\hat u_k \hat v_k (\hat{\bar w} - w_k)},
\end{aligned}
\end{equation}
and
\begin{multline}\label{eq:mfp_sign_threshold}
sign \frac{\partial \hat{\bar w}}{\partial \tau} = 
sign \sum_{k \in K}{\hat u_k \hat v_k (\hat{\bar w} - w_k)} = \\ 
sign \Big(\hat{\bar w} - \frac{\sum_{k \in K}{\hat u_k \hat v_k w_k}}{\sum_{k \in K}{\hat u_k \hat v_k}}\Big).
\end{multline}
\end{corollary}

@@

An interesting consequence of Eq.~\ref{eq:mfp_intermediate} is
\begin{equation}
\frac{\partial \hat{\bar w}}{\partial C_k} = 
\hat v_k w_k (\hat u\tr[M]_k - \hat u_k) = 
w_k \Big(\sum_{j=1}^{n}{\hat u_j M_{j,k} \hat v_k} - \hat u_k \hat v_k\Big).
\end{equation}
Here, $\hat u_k$ can be regarded as the \emph{reproductive value} of type $k$~\citep[pg.~27]{Fisher1930}, which gives the relative contribution of type $k$ to the long-term population (see~\nameref{sec:AppB}).
Consequently, $\hat u_k \hat v_k$ is the \emph{ancestor frequency} of type $k$~\citep{Hermisson2002}, namely the fraction of the equilibrium population descended from type $k$.
The sum $\sum_{j=1}^{n}{\hat u_j M_{j,k} \hat v_k}$ can be similarly interpreted as the fraction of the equilibrium population descended from individuals that transitioned from type $k$ to another type (via the $k$ column of the transition matrix $M$), conditioned on transition occurring.

@@

Corollary~\ref{cor:rvp} sheds light on why we require $M$ to be primitive.
If $M$ is primitive then individuals of type $k$ can transition into any other type in a finite number of generations.
So individuals with below-average fitness can have descendants with above-average fitness, and increased generation of variation in these individuals will increase the stable population mean fitness.
In contrast, if $M$ is not primitive, individuals with below-average fitness are "doomed" and increasing the generation of variation in these individuals can only hasten their removal from the population, affecting the rate at which the population converges to the stable equilibrium, but not necessarily the stable population mean fitness.
For example, if we set $\beta=0$ in \nameref{sec:mut_model_2}, $M$ becomes triangular and imprimitive, and the stable mean fitness becomes $\bar{w} = (1-\delta C_0)w_0$, which is not affected by changes in $C_k$ for $k \ge 1$~\citep[see also][Fig.~1A]{Agrawal2002,Ram2012}.

@@

\paragraph*{Modifier model.}
Consider the case of two modifier alleles, $m$ and $M$, inducing different transition probabilities $C=\diag{C_1, \ldots, C_n}$ and $\tilde{C}=\diag{\tilde{C}_1, \ldots, \tilde{C}_n}$, respectively.
The frequencies of type $A_k$ linked to modifier $m$ or $M$ are $f_k$ and $g_k$, respectively, where $\sum_{k=1}^{n}{(f_k + g_k)}=1$.
$\bar{w}$ now ensures that $\sum_{k=1}^{n}{(f_k' + g_k')}=1$, 
and the rest of the model parameters are the same as in Eq.~\ref{eq:model}.

The frequencies in the next generation, $f'$ for allele $m$ and $g'$ for allele $M$, are given by
\begin{equation}
\begin{cases}
\bar{w} f' &= (I-C+MC)D f \\
\bar{w} g' &= (I-\tilde{C}+M\tilde{C})D g
\end{cases}.
\label{eq:modifier_model}
\end{equation}
Here, $\bar{w}=\sum_{k=1}^{n}{(f_k + g_k) w_k}$ is the mean fitness of the entire population. 
Note that Eq.~\ref{eq:model} is the special case of Eq.~\ref{eq:modifier_model} where allele $M$ is absent, i.e. $g_k=0$ for all $k$.

Result~\ref{result:mfp} provides a condition under which increasing the transition rate $C_k$ from type $A_k$ will increase the stable population mean fitness.
Could a modifier allele that increases $C_k$ increase in frequency when initially rare in the population?
To answer this we analyze the stability of resident modifier allele $m$ with transition rates $C_k$ to  invasion by a modifier allele $M$ with rates $\tilde{C}_k$ under Eqs.~\ref{eq:modifier_model}.

The equilibrium of Eqs.~\ref{eq:modifier_model} when modifier allele $M$ is absent from the population is $(\hat v,0)$, where $\hat v$ is given in Eq.~\ref{eq:model_equilibrium} and $g_k=0$ for all $k$. 
The stability of allele $m$ to invasion by allele $M$ is determined by the leading eigenvalue $\lambda_1$ of $\cl_{ex}$ the external stability matrix of the equilibrium $(\hat v,0)$, which, in turn, is determined by the Jacobian $\cj$ of the system in Eqs.~\ref{eq:modifier_model} evaluated at the equilibrium $(\hat v,0)$, where
\begin{equation}
\cj = \begin{pmatrix}\cl_{in} & 0 \\ 0 & \cl_{ex} \end{pmatrix},
\end{equation}
and $\cl_{in}$ is the local stability matrix of the equilibrium $(\hat v,0)$ in the space $\sum_{k=1}^n{f_k}=1$.
The zero block matrices are due to the complete linkage between the modifier and the types $A_k$ and to the lack of transition (i.e., mutation) between the modifier alleles.

$\cl_{ex}$ can be written as
\begin{equation}\label{eq:Lex}
\cl_{ex} = (I - \tilde{C} + M \tilde{C}) \tilde{D} ,
\end{equation}
where $\tilde{D} =D/\hat{\bar{w}}$ is the diagonal matrix with entries $w_k/\hat{\bar{w}}$ for all $k$, and $\hat{\bar{w}}$ is the stable population mean fitness in the absence of the modifier allele $M$.
$\lambda_1$, the leading eigenvalue of $\cl_{ex}$, coincides with $\tilde{\bar{w}}$, the maximal mean fitness associated with Eq.~\ref{eq:Lex}.
Thus we can apply Result~\ref{result:mfp} to $\lambda_1$ and obtain the following result.

\begin{result}[Evolution of increased genetic variation]
Let $\lambda_1$ be the the leading eigenvalue of the external stability matrix $\cl_{ex}$.
If the transition rates induced by the modifier alleles $m$ and $M$ are equal, i.e., $\tilde{C}_k=C_k$ for all $k$, then 
\begin{equation}
\lambda_1=1,
\end{equation}
and
\begin{equation}
sign \frac{\partial \lambda_1}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
sign(\hat{\bar w} - w_k).
\end{equation}

Therefore, an initially rare modifier allele $M$ with transition rates slightly different from the resident allele $m$ can successfully invade the population ($\lambda_1>1$) if $M$ increases the probability of transition from types with below-average fitness, thereby increasing the  mean fitness $\tilde{\bar w}$.
\end{result}

@@

Now, applying Result~\ref{result:mfp} (eq.~\ref{eq:mfp_sign}) to Eq.~\ref{eq:Lex}, the sign of the derivative of $\lambda_1$ with respect to $\tilde{C_k}$ is
\begin{equation}
sign \frac{\partial \lambda_1}{\partial \tilde{C}_k} =
sign\left(\lambda_1 - \frac{w_k}{\hat{\bar w}}\right).
\end{equation}
Thus
\begin{multline}
sign \left.\frac{\partial \lambda_1}{\partial \tilde{C}_k}\right\rvert_{\tilde{C}_k = C_k} = 
sign\left(1 - \frac{w_k}{\hat{\bar w}}\right) =  
sign(\hat{\bar w} - w_k),
\end{multline}
since $\hat{\bar w}>0$.
This completes the proof.
\end{proof}

@@

\bibliographystyle{agsm}
%\bibliography{/Users/yoavram/Documents/library}
\bibliography{ms}

>>>>

\documentclass[12pt, twocolumn]{extarticle}
\usepackage{geometry}
\geometry{
a4paper,
total={170mm,257mm},
left=20mm,
top=20mm,
headheight=12pt
}

%\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx} % Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
% TeX will automatically convert eps --> pdf in pdflatex

@@

@@

% Title page
\title{Generation of Variation and a Modified Mean Fitness Principle: \\ Necessity is the Mother of Genetic Invention}
%\title{Conditional Generation of Variation and Mean Fitness Increase: \\ Necessity is the Mother of Genetic Invention}
% \title{A General Principle for the Evolution of Fitness-Associated Generation of Variation}

@@

Another departure from the \emph{linear variation} assumption of the \emph{reduction principle} for mutation rates involves a mechanism by which the mutation rate increases in individuals of low fitness -- a mechanism first observed in stressed bacteria~\citep{Foster2007}, although not in a constant environment. A modifier gene that changes transition rates in only a subset of types exemplifies the \emph{principle of partial control} \citep{Altenberg1984,Altenberg2012a} a general statement about non-linear variation, which proposes that, ``When the modifier gene has only partial control over the
transformations occurring at loci under selection, then it may be possible for the part which it controls to evolve an increase in rates.''

\citet{Ram2012} demonstrated that even in a constant environment, increasing the mutation rate of individuals with below-average fitness increases the population mean fitness, rather than decreases it.
Their analysis assumed  infinite population size and fitness  determined by the number of mutant alleles accumulated in the genotype.
In their models, the only departure from the \emph{reduction principle} conditions was the unequal scaling of mutation probabilities between different genotypes introduced by the correlation between the mutation rate and fitness.
A similar result has been demonstrated for conditional dispersal~\citep[Th.~39]{Altenberg2012b}, fitness-associated recombination~\citep{Hadany2003a} and for condition-dependent sexual reproduction~\citep{Hadany2007a}, and empirical evidence suggests these mechanisms are common in nature~\citep{Ram2016}.

@@

\paragraph*{General model.}\label{sec:general_model}
Consider a large  population with an arbitrary set of types $A_1, A_2, \ldots, A_n$.
The frequency and fitness of individuals of type $A_k$ are $f_k$ and $w_k$, respectively.
The probability that an individual of type $A_k$ will transition to some other type is $C_k$, and given a transition occurs, the probability that it will transition to type $A_j$ is $M_{j,k}$.
Therefore, the change in the frequencies of type $A_k$ is described by the transformation $f \to f'$: 
\begin{equation}
\bar{w} f'_k = (1-C_k) w_k f_k + \sum_{j=1}^{n}{C_j M_{k,j} w_j f_j},
\label{eq:model_sum}
\end{equation}
or in matrix form (introduced to describe conditional dispersal in \citet{Altenberg2012a})
\begin{equation}
\bar{w} f' = (\ci - C + MC)D f,
\label{eq:model}
\end{equation}
where $f=(f_1, f_2, \ldots, f_n)$ is a frequency vector with $f_k\ge 0$ and  $\sum_{k=1}^n{f_k} = 1$;
$D$ is a positive diagonal matrix with entries $w_k$ such that $w_k \ne w_j$ for some $k \ne j$; 
$C$ is a positive diagonal  matrix with entries $C_k$, where $C_k<1$ for at least one $k$;
$M$ is an irreducible  column-stochastic matrix: $M_{j,k} \ge 0$ for all $j,k$, $\sum_{j=1}^n {M_{j,k}} = 1$ for all $k$, and $(M^{\ell})_{j,k} > 0$ for all $j,k$ for some positive integer $\ell$ (the latter is the condition for primitivity~\citep[see][Appendix 3]{Otto2007});
$\ci$ is the $n\times n$ identity matrix;
and $\bar{w}$ is the normalizing factor such that $\sum_{k=1}^n{f'_k}=1$ and is equal to the population mean fitness $\bar w=\sum_{k=1}^n f_kw_k$.

@@

Transition between types is determined by a combination of two effects:
(i) the probability of transitioning \emph{out} of type $A_k$ is determined by $C_k$;
(ii) given a transition out of type $A_k$, the distribution of the destination types $A_i$ is given by $M_{i,k}$ (note the index order).
Importantly, different types can have different rates.
That is, $C_i \ne C_j$ for some $i,j$. The case $C_i = C_j$ for all $i,j$ is covered by the \emph{reduction principle}~(see \citealt{Altenberg1984,Altenberg1987,Altenberg2009,Altenberg2017}).

@@

This model is a special case of the general model (Eq.~\ref{eq:model}) where
\begin{equation} 
{M} = \frac{1}{n-1} \begin{bmatrix}
0 & 1 & 1 & \ldots & 1 \\
1 & 0 & 1 & \ldots & \vdots \\
1 & 1 & 0 & \ldots & \vdots \\
\vdots & \vdots & \vdots & \ddots & \vdots \\
1 & \ldots & \ldots & \ldots & 0
\end{bmatrix},
\end{equation}
with zeros on the diagonal and $\frac{1}{n-1}$ elsewhere.
Note that here ${M}$ is primitive (i.e., irreducible and aperiodic).% Mutation model 2
\paragraph*{Mutation model 2.}\label{sec:mut_model_2}

@@

Therefore, setting 
\begin{equation}
{M} = \begin{bmatrix}
1-\delta & \beta & 0 &  \cdots & 0\\
\delta & 1-\delta-\beta & \beta &  \ddots & 0\\
0 & \delta & 1-\delta-\beta & \ddots & 0 \\
\vdots & \ddots & \ddots & \ddots & \beta \\
0 & 0 & 0 & \delta & 1-\beta
\end{bmatrix},
\end{equation}
with the beneficial, neutral, and deleterious mutation probabilities on the three main diagonals and zeros elsewhere, 
Eq.~\ref{eq:mutation_model_2} can be viewed as a special case of Eq.~\ref{eq:model}.
Here, too, ${M}$ is primitive as long as $\delta, \beta > 0$.

@@

After reproduction, offspring of individuals from deme $A_k$ return to their parental deme with probability $1-C_k$, or migrate to a different deme $A_j$ with probability $C_k M_{j,k}$, where the matrix $M$ is irreducible, $C_k > 0$, $C\ne I$, $M_{j,k} \ge 0$, and $\sum_{j=1}^{n}{M_{j,k}} = 1$ for all $k=1, \ldots, n$.
Therefore, $1-C_k$ can be interpreted as a \emph{homing rate}. When the homing rate differs for different demes, it has been called \emph{conditional dispersal} \citep{McPeek1992,Kirkland2006}.

@@

When $C_k = \tau$ for all $k$ and fitnesses are not all equal, then a culturally transmitted variant in $\tau$ will spread if and only if $\tau$ is reduced, the first manifestation of the reduction principle shown in cultural evolution \citep[pp. 203--206]{Altenberg1984}.

In a more specific example, for the case of \emph{symmetric individual learning}~\citep{Borenstein2008}, learning is parameterized by its breadth of exploration $b$ and the mature phenotype $j$ is randomly drawn from $2b+1$ phenotypes symmetrically and uniformly distributed around the initial phenotype $k$, with the limitation that any "spillover" of phenotypes below $1$ or above $n$ is "absorbed" by those boundaries.
This "absorption" ensures $M$ is column-stochastic. 
In other words, given initial phenotype $k$, the probability of maturation to phenotype $j$, where $k-b \le j \le k+b$ is $1/(2b+1)$, but any phenotype $j<1$ actually becomes $j=1$ and any phenotype $j>n$ actually becomes $j=n$.
The probability for maturation to other phenotypes is $0$.

@@

% Mean fitness principle
\paragraph*{A modified mean-fitness principle.}

We first focus on the stable population mean fitness.
We  show that if the transition rate from types with below-average fitness increases, then the stable population mean fitness increases, too.  We refer to this as a ``modified'' mean fitness principle (MMFP) because for modifiers of mutation and migration rates the standard mean fitness principle involves a decrease in the mean fitness with increasing rates.

@@

The following result constitutes an MMFP for the
direction of the sensitivity of the equilibrium mean fitness $\hat{\bar w}$ to changes
in $C_k$, the probability of transition from $A_k$.
\medskip

\begin{result}[Modified mean fitness principle]\label{result:mfp}
Let $\hat{\bar w}$ be the leading eigenvalue of $(I-C+MC)D$,
and $\hat u$ and $\hat v$ be the corresponding positive left and right eigenvectors, such that $\sum_{k=1}^{n}\hat v_k=1$ and $\sum_{k=1}^{n}\hat u_k\hat v_k=1$.
Then,
\begin{equation}\label{eq:mfp}
\frac{\partial \hat{\bar w}}{\partial C_k} = 
\frac{\hat u_k \hat v_k}{C_k} (\hat{\bar w} - w_k),
\end{equation}
or in simpler terms,
\begin{equation}\label{eq:mfp_sign}
sign\frac{\partial \hat{\bar w}}{\partial C_k} = 
sign(\hat{\bar w} - w_k).
\end{equation}

@@

\begin{proof}
Following the analysis in \citet [Th.~39]{Altenberg2012a}, we apply to \eqref {eq:model_equilibrium} the formula in~\citet{Caswell1978} (see eq.~\ref{eq:Caswells_formula} in~\nameref{sec:AppA}),
\begin{equation}
\frac{\partial \hat{\bar w}}{\partial C_k} = 
\hat u\tr \frac{\partial (I-C+MC)D}{\partial C_k} \hat v.
\end{equation}

@@

Then,
\begin{multline}\label{eq:mfp_intermediate}
%\begin{aligned}
\hat u\tr \frac{\partial (I-C+MC)D}{\partial C_k} \hat v = 
\hat u\tr (0 - Z_k + M Z_k)D \hat v = \\
-\hat v_k \hat u_k w_k + \hat v_k w_k \hat u\tr[M]_k = \\
\hat v_k w_k (\hat u\tr[M]_k - \hat u_k). 
%\end{aligned}
\end{multline}

@@

% Multiple types
\paragraph*{}
We turn our attention to the case where the transition rates from a subset $K$ of the types are equal, that is, $C_j \equiv C_i$ for $i,j \in K$.
In this case, Eq.~\ref{eq:mfp} leads directly to the following by setting $\tau = C_i$ for $i\in K$.

\begin{corollary}
The sensitivity of the stable population mean fitness to change in the rate of transition $\tau$ from types in $K$ is
\begin{multline}\label{eq:mfp_K}
%\begin{aligned}
\frac{\partial \hat{\bar w}}{\partial \tau} = 
\hat u\tr \Big( \sum_{k \in K}{\frac{\partial (I - C + MC)D}{\partial \tau}} \Big) \hat v = \\
\frac{1}{\tau} \sum_{k \in K}{\hat u_k \hat v_k (\hat{\bar w} - w_k)},
%\end{aligned}
\end{multline}
and
\begin{multline}\label{eq:mfp_sign_threshold}
sign \frac{\partial \hat{\bar w}}{\partial \tau} = 
sign \sum_{k \in K}{\hat u_k \hat v_k (\hat{\bar w} - w_k)} = \\ 
sign \Big(\hat{\bar w} - \frac{\sum_{k \in K}{\hat u_k \hat v_k w_k}}{\sum_{k \in K}{\hat u_k \hat v_k}}\Big).
\end{multline}
\end{corollary}

@@

An interesting consequence of Eq.~\ref{eq:mfp_intermediate} observed in \citet [Th.~39]{Altenberg2012a} is
\begin{equation}
\begin{aligned}
\frac{\partial \hat{\bar w}}{\partial C_k} = 
\hat v_k w_k (\hat u\tr[M]_k - \hat u_k) = \\
w_k \Big(\sum_{j=1}^{n}{\hat u_j M_{j,k} \hat v_k} - \hat u_k \hat v_k\Big).
\end{aligned}
\end{equation}
Here, $\hat u_k$ can be regarded as the \emph{reproductive value} of type $k$~\citep[pg.~27]{Fisher1930}, which gives the relative contribution of type $k$ to the long-term population (see~\nameref{sec:AppB}).
Consequently, $\hat u_k \hat v_k$ is what ~\citet{Hermisson2002} call the \emph{ancestor frequency} of type $k$, which is better described as the fraction of the population in the distant future that descended from the individuals of type $k$ present at a particular moment in a population at equilibrium.
The sum $\sum_{j=1}^{n}{\hat u_j M_{j,k} \hat v_k}$ can be similarly interpreted as the fraction of the equilibrium population descended from individuals that transitioned from type $k$ to another type (via the $k$ column of the transition matrix $M$), conditioned on transition occurring.

@@

Corollary~\ref{cor:rvp} sheds light on why we require $M$ to be irreducible.
If $M$ is irreducible then individuals of type $k$ can transition into any other type in a finite number of generations.
So individuals with below-average fitness can have descendants with above-average fitness, and increased generation of variation in these individuals will increase the stable population mean fitness.
In contrast, if $M$ is not irreducible, it is possible that some individuals with below-average fitness are ``doomed'' since they cannot produce descendants with above-average fitness, so increasing the generation of variation in these individuals can only hasten their removal from the population, affecting the rate at which the population converges to the stable equilibrium, but not necessarily the stable population mean fitness.
For example, if we set $\beta=0$ in \nameref{sec:mut_model_2}, $M$ becomes triangular and reducible, and the stable mean fitness becomes $\bar{w} = (1-\delta C_0)w_0$, which is not affected by changes in $C_k$ for $k \ge 1$~\citep[see also][Fig.~1A]{Agrawal2002,Ram2012}.

@@

\paragraph*{Modifier model.}
Consider the case of two modifier alleles, $b$ and $B$, inducing different transition probabilities $C=\diag{C_1, \ldots, C_n}$ and $\tilde{C}=\diag{\tilde{C}_1, \ldots, \tilde{C}_n}$, respectively.
The frequencies of type $A_k$ linked to modifiers $b$ or $B$ are $f_k$ and $g_k$, respectively, where $\sum_{k=1}^{n}{(f_k + g_k)}=1$.
$\bar{w}$ now ensures that $\sum_{k=1}^{n}{(f_k' + g_k')}=1$, 
and the rest of the model parameters are the same as in Eq.~\ref{eq:model}.

The frequencies in the next generation, $f'$ for allele $b$ and $g'$ for allele $B$, are given by
\begin{equation}
\begin{cases}
\bar{w} f' &= (I-C+MC)D f \\
\bar{w} g' &= (I-\tilde{C}+M\tilde{C})D g
\end{cases}.
\label{eq:modifier_model}
\end{equation}
Here, $\bar{w}=\sum_{k=1}^{n}{(f_k + g_k) w_k}$ is the mean fitness of the entire population. 
Note that Eq.~\ref{eq:model} is the special case of Eq.~\ref{eq:modifier_model} where allele $M$ is absent, i.e. $g_k=0$ for all $k$.

Result~\ref{result:mfp} provides a condition under which increasing the transition rate $C_k$ from type $A_k$ will increase the stable population mean fitness.
Could a modifier allele that increases $C_k$ increase in frequency when initially rare in the population?
To answer this we analyze the stability of resident modifier allele $b$ with transition rates $C_k$ to  invasion by a modifier allele $B$ with rates $\tilde{C}_k$ under Eqs.~\ref{eq:modifier_model}.

The equilibrium of Eqs.~\ref{eq:modifier_model} when modifier allele $B$ is absent from the population is $(\hat v,0)$, where $\hat v$ is given in Eq.~\ref{eq:model_equilibrium} and $g_k=0$ for all $k$. 
The stability of allele $b$ to invasion by allele $B$ is determined by the leading eigenvalue $\lambda_1$ of $\cl_{ex}$ the external stability matrix of the equilibrium $(\hat v,0)$, which, in turn, is determined by the Jacobian $\cj$ of the system in Eqs.~\ref{eq:modifier_model} evaluated at the equilibrium $(\hat v,0)$, where
\begin{equation}
\cj = \begin{pmatrix}\cl_{in} & 0 \\ 0 & \cl_{ex} \end{pmatrix},
\end{equation}
and $\cl_{in}$ is the local stability matrix of the equilibrium $(\hat v,0)$ in the space $\sum_{k=1}^n{f_k}=1$.
The zero block matrices are due to the complete linkage between the modifier and the types $A_k$ and to the lack of transition (i.e., mutation) between the modifier alleles.

$\cl_{ex}$ can be written as
\begin{equation}\label{eq:Lex}
\cl_{ex} = (I - \tilde{C} + M \tilde{C}) \tilde{D} ,
\end{equation}
where $\tilde{D} =D/\hat{\bar{w}}$ is the diagonal matrix with entries $w_k/\hat{\bar{w}}$ for all $k$, and $\hat{\bar{w}}$ is the stable population mean fitness in the absence of the modifier allele $B$.
$\lambda_1$, the leading eigenvalue of $\cl_{ex}$, coincides with $\tilde{\bar{w}}$, the maximal mean fitness associated with Eq.~\ref{eq:Lex}.
Thus we can apply Result~\ref{result:mfp} to $\lambda_1$ and obtain the following result.

\begin{result}[Evolution of increased genetic variation]
Let $\lambda_1$ be the leading eigenvalue of the external stability matrix $\cl_{ex}$.
If the transition rates induced by the modifier alleles $b$ and $B$ are equal, i.e., $\tilde{C}_k=C_k$ for all $k$, then 
\begin{equation}
\lambda_1=1,
\end{equation}
and
\begin{equation}
sign \frac{\partial \lambda_1}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
sign(\hat{\bar w} - w_k).
\end{equation}

Therefore, an initially rare modifier allele $B$ with transition rates slightly different from the resident allele $m$ can successfully invade the population ($\lambda_1>1$) if $B$ increases the probability of transition from types with below-average fitness, thereby increasing the  mean fitness $\tilde{\bar w}$.
\end{result}

@@

Now, applying Result~\ref{result:mfp} (eq.~\ref{eq:mfp_sign}) to Eq.~\ref{eq:Lex}, the sign of the derivative of $\lambda_1$ with respect to $\tilde{C_k}$ is
\begin{equation}
sign \frac{\partial \lambda_1}{\partial \tilde{C}_k} =
sign\left(\lambda_1 - \frac{w_k}{\hat{\bar w}}\right).
\end{equation}
Thus
\begin{equation}
\begin{aligned}
sign \left.\frac{\partial \lambda_1}{\partial \tilde{C}_k}\right\rvert_{\tilde{C}_k = C_k} = 
sign\left(1 - \frac{w_k}{\hat{\bar w}}\right) = \\
sign(\hat{\bar w} - w_k),
\end{aligned}
\end{equation}
since $\hat{\bar w}>0$.
This completes the proof.
\end{proof}

@@

\bibliographystyle{agsm}
%\bibliography{/Users/yoavram/Documents/library}
\bibliography{ms+Lee2}
